The truncated KdV framework for modeling anomalous waves induced by abrupt depth changes

### 2019-07-17 WavesData-2019/LatexNotes/notes.tex
NM

% Basic physical parameters and scales
\newcommand{\freqp}{f_p}
\newcommand{\amp}{a}
\newcommand{\ampscale}{A}
%\newcommand{\etastd}{\eta_{\text std}}
% Parameters that change crossing the ADC
\newcommand{\depth}{d}
\newcommand{\dup}{\depth_{-}}
\newcommand{\ddn}{\depth_{+}}
\newcommand{\dupdn}{\depth_{\pm}}
\newcommand{\lam}{\lambda}
\newcommand{\lamup}{\lam_{-}}
\newcommand{\lamdn}{\lam_{+}}
\newcommand{\lamupdn}{\lam_{\pm}}
\newcommand{\lamfac}{N}
\newcommand{\drat}{D}
\newcommand{\dratdn}{\drat_+}
\newcommand{\dratupdn}{\drat_{\pm}}
% Statistical quantities
\newcommand{\skw}{\text{skew}}
\newcommand{\var}{\text{var}}
\newcommand{\En}{\mathcal{E}}
% Dimensionless parameters
\newcommand{\epsup}{\eps_0}
\newcommand{\delup}{\delta_0}

\newcommand{\ampp}{\mathcal{A}}
% Hamiltonian and Gibbs stuff
\newcommand{\Ham}{H}
\newcommand{\Hup}{\Ham^{-}}
\newcommand{\Hdn}{\Ham^{+}}
\newcommand{\Gibbs}{\mathcal{G}}
\newcommand{\Gup}{\Gibbs^{-}}
\newcommand{\Gdn}{\Gibbs^{+}}
\newcommand{\meanup}[1]{\mean{#1}_{-}}
\newcommand{\meandn}[1]{\mean{#1}_{+}}

@@

Dropping the tildes, the dimensionless KdV equation takes the form
\begin{equation}
\label{dimlessKdV1}
u_t + \frac{3}{2} \pi \epsup \drat^{-1} \, u u_x + \frac{1}{6} \pi^3 \lamfac^{-2} \delup^2 \drat \, u_{xxx} = 0
\qquad \text{for } x \in [-\pi,\pi]
\end{equation}
where dimensionless variables are as follows
\begin{align}
&\epsup = \amp/\dup \, , \\
&\delup = \dup/\lamup = \sqrt{\dup \freqp^2/g} \, \\
&\drat = {\depth}/{\dup} \, .
\end{align}
Here, $\drat = \dratupdn$ is the local dimensionless depth, which changes value crossing the ADC. In particular, since \eqref{dimlessKdV1} is in the reference frame that moves with the characteristic wave speed, we can assume that the ADC is met at $t = T_{ADC} = 0$. Then we have
\begin{equation}
\drat = 
\begin{cases}
1 		&\quad \mbox{for } t<0 \\
\dratdn 	&\quad \mbox{for } t>0
\end{cases}
\end{equation}
where $\dratdn = \ddn/\dup$.
See Table \ref{paramtable} for a summary of important parameters.

@@

Therefore, lets stick with the original normalization \eqref{dimlessKdV1}, repeated here
\begin{equation}
\label{dimlessKdV2}
u_t + \frac{3}{2} \pi \epsup \drat^{-1} \, u u_x + \frac{1}{6} \pi^3 \lamfac^{-2} \delup^2 \drat \, u_{xxx} = 0
\qquad \text{for } x \in [-\pi,\pi]
\end{equation}
If we need to draw a comparison to the PNAS paper, we set something like $\epsup = E^{1/2}L^{-3/2}$ and $\delup^2 = L^{-3}$, etc.

@@

We now consider the Hamiltonian structure of the dimensionless variable-depth KdV equation \eqref{dimlessKdV1}, over dimensionless domain $x \in [-\pi,\pi]$. First, the spatial average of any quantity is given by
\begin{equation}
\mean{q} = \frac{1}{2\pi} \int_{-\pi}^{\pi} q \dx \, .
\end{equation}
Appealing to near Ergodicity, we will typically assume equivalence between spatial, temporal, and ensemble averages.

We measure the state variable $u$ as the displacement of the surface from equilibrium, so that $\mean{u} = 0$. Consequently, the momentum of $u$ vanishes
\begin{equation}
M[u] \defeq \int_{-\pi}^{\pi} u \dx = 0
\end{equation}
Next, the definition of the characteristic amplitude in \eqref{ampdefn} implies that $\mean{u^2} = 1$. Consequently, the energy of $u$ is fixed as
\begin{equation}
E[u] \defeq \frac{1}{2} \int_{-\pi}^{\pi} u^2 \dx = \pi
\end{equation}
TO DO: Add Fourier stuff including Parseval.
\vsp{4}

@@

Then the Hamiltonian of \eqref{dimlessKdV1} is given by
\begin{equation}
\Ham = 3 \pi \epsup \drat^{-1} \, H_3 - \frac{4 \pi^3}{3 \lamfac^2} \delup^2 \drat \, H_2
\end{equation}
where $\drat$ changes value across the ADC. More specifically, there exists an upstream and a downstream Hamiltonian
\begin{align}
&\Hup = 3 \pi \epsup \, H_3 - \frac{4 \pi^3}{3 \lamfac^2} \delup^2 \, H_2 \\
&\Hdn = 3 \pi \epsup \dratdn^{-1} \, H_3 - \frac{4 \pi^3}{3 \lamfac^2} \delup^2 \dratdn \, H_2
\end{align}
$\Hup$ is a conserved quantity for $t<0$ and $\Hdn$ is a conserved quantity for $t>0$. Each Hamiltonian $\Ham^{\pm}$ generates a corresponding Gibbs measure $\Gibbs^{\pm}$ with corresponding ensemble means $\mean{}_{\pm}$.

(FILL IN DEFINITIONS)

@@

The statistical matching condition is
\begin{equation}
\meanup{\Hdn} = \meandn{\Hdn}
\end{equation}
An experimental observation is that the incoming skewness is small, thus
\begin{equation}
\meanup{H_3} \approx 0
\end{equation}
which immediately gives
\begin{equation}
3 \pi \epsup \dratdn^{-1} \meandn{H_3} = \frac{4 \pi^3}{3 \lamfac^2} \delup^2 \dratdn \left( \meandn{H_2} - \meanup{H_2} \right)
\end{equation}
or more simply
\begin{equation}
\label{H3H2}
\meandn{H_3} = \frac{4 \pi^2}{9 \lamfac^2} \frac{\delup^2}{\epsup} \dratdn^2 \,\, \Delta \mean{H_2}
\end{equation}
where $\Delta \mean{H_2} =  \meandn{H_2} - \meanup{H_2}$  indicates the difference between the upstream and downstream values.
To convert to dimensional, i.e.~experimental, values we use
\begin{align}
&\mean{u^3} = \frac{\mean{\eta^3}}{a^3} = 
\frac{\mean{\eta^3}}{\mean{\eta^2}^{3/2}} = \skw(\eta) \\
&\mean{u_x^2} = \left(\frac{\lam}{a} \right)^2 \mean{\eta_x^2} 
= \left(\frac{\lam}{a} \right)^2 \var(\eta_x)
\end{align}
Then \eqref{H3H2} gives the relationship
\begin{equation}
\frac{\skw(\eta)}{\Delta \var(\eta_x)} =
3 \mu D^2 \left( \frac{h_0}{a} \right)^3 = 3 \eps^{-3} \mu D^2
\end{equation}
In particular, the ratio on the left, which can be measured in the experiments, is predicted to scale as the inverse cube of the wave amplitude (as well as the square of the depth ratio).

>>>>

% Basic physical parameters and scales
\newcommand{\freqp}{f_p}
\newcommand{\amp}{a}
\newcommand{\ampscale}{A}
%\newcommand{\etastd}{\eta_{\text std}}
% Parameters that change crossing the ADC
\newcommand{\depth}{d}
\newcommand{\dup}{\depth_{-}}
\newcommand{\ddn}{\depth_{+}}
\newcommand{\dupdn}{\depth_{\pm}}
\newcommand{\lam}{\lambda}
\newcommand{\lamup}{\lam_{-}}
\newcommand{\lamdn}{\lam_{+}}
\newcommand{\lamupdn}{\lam_{\pm}}
\newcommand{\lamfac}{N}
\newcommand{\drat}{D}
\newcommand{\dratdn}{\drat_+}
\newcommand{\dratupdn}{\drat_{\pm}}
% Statistical quantities
\newcommand{\skw}{\text{skew}}
\newcommand{\var}{\text{var}}
\newcommand{\En}{\mathcal{E}}
\newcommand{\Mo}{\mathcal{M}}
% Dimensionless parameters
\newcommand{\epsup}{\eps_0}
\newcommand{\delup}{\delta_0}
% Fourier
\newcommand{\uhat}{\hat{u}}

\newcommand{\ampp}{\mathcal{A}}
% Hamiltonian and Gibbs stuff
\newcommand{\Ham}{H}
\newcommand{\Hup}{\Ham^{-}}
\newcommand{\Hdn}{\Ham^{+}}
\newcommand{\Hupdn}{\Ham^{\pm}}
\newcommand{\Gibbs}{\mathcal{G}}
\newcommand{\Gup}{\Gibbs^{-}}
\newcommand{\Gdn}{\Gibbs^{+}}
\newcommand{\Gupdn}{\Gibbs^{\pm}}
\newcommand{\thup}{\theta^{-}}
\newcommand{\thdn}{\theta^{+}}
\newcommand{\thupdn}{\theta^{\pm}}
\newcommand{\meanup}[1]{\mean{#1}_{-}}
\newcommand{\meandn}[1]{\mean{#1}_{+}}

@@

Dropping the tildes, the dimensionless KdV equation takes the form
\begin{equation}
\label{dimlessKdV1}
u_t + \frac{3 \pi}{2} \epsup \drat^{-1} \, u u_x + \frac{\pi^3}{6 \lamfac^2} \delup^2 \drat \, u_{xxx} = 0
\qquad \text{for } x \in [-\pi,\pi]
\end{equation}
where dimensionless variables are as follows
\begin{align}
&\epsup = \amp/\dup \, , \\
&\delup = \dup/\lamup = \sqrt{\dup \freqp^2/g} \, \\
&\drat = {\depth}/{\dup} \, .
\end{align}
Here, $\drat = \dratupdn$ is the local dimensionless depth, which changes value crossing the ADC. In particular, since \eqref{dimlessKdV1} is in the reference frame that moves with the characteristic wave speed, we can assume that the ADC is met at $t = T_{ADC} = 0$. Then we have
\begin{equation}
\drat = 
\begin{cases}
1 		&\quad \mbox{for } t<0 \\
\dratdn 	&\quad \mbox{for } t>0
\end{cases}
\end{equation}
where $\dratdn = \ddn/\dup$.
See Table \ref{paramtable} for a summary of important parameters.

@@

Therefore, lets stick with the original normalization \eqref{dimlessKdV1}, recast here in different form
\begin{align}
\label{dimlessKdV2}
&u_t + C_3 \drat^{-1} \, u u_x + C_2 \drat \, u_{xxx} = 0
\qquad \text{for } x \in [-\pi,\pi] \\
&C_3 = \frac{3}{2} \pi \epsup  \\
&C_2 = \frac{\pi^3 \delup^2}{6 \lamfac^2} 
\end{align}
If we need to draw a comparison to the PNAS paper, we set something like $\epsup = E^{1/2}L^{-3/2}$ and $\delup^2 = L^{-3}$, etc.

@@

We now consider the Hamiltonian structure of the dimensionless variable-depth KdV equation \eqref{dimlessKdV2}, over dimensionless domain $x \in [-\pi,\pi]$. First, the spatial average of any quantity is given by
\begin{equation}
\mean{q} = \frac{1}{2\pi} \int_{-\pi}^{\pi} q \dx \, .
\end{equation}
Appealing to near Ergodicity, we will typically assume equivalence between spatial, temporal, and ensemble averages.

We measure the state variable $u$ as the displacement of the surface from equilibrium, so that $\mean{u} = 0$. Consequently, the momentum of $u$ vanishes
\begin{equation}
\Mo[u] \defeq \int_{-\pi}^{\pi} u \dx = 0
\end{equation}
Next, the definition of the characteristic amplitude in \eqref{ampdefn} implies that $\mean{u^2} = 1$. Consequently, the energy of $u$ is fixed as
\begin{equation}
\En[u] \defeq \frac{1}{2} \int_{-\pi}^{\pi} u^2 \dx = \pi
\end{equation}

We introduce the Fourier series
\begin{equation}
u = \sum_{k=-\Lambda}^{\Lambda} \uhat_k e^{i k x}
\end{equation}
where
\begin{equation}
\uhat_k = \frac{1}{2 \pi} \int_{-\pi}^{\pi} u(x) e^{-i k x} \dx
\end{equation}
The above sum can be infinite $\Lambda = \infty$ or truncated at a finite wavenumber $\Lambda < \infty$.
Further, since $u$ is real valued, we have $\uhat_{-k} = {\uhat_{k}}^*$.
By Parseval's identity, we have
\begin{equation}
\En[u] = \pi = \frac{1}{2} \int u^2 \dx = \pi \sum_{k=-\Lambda}^{\Lambda} \abs{\uhat_k}^2 = 2 \pi \sum_{k=1}^{\Lambda} \abs{\uhat_k}^2
\end{equation}
% The above agrees with the first line on top of p. 4 in Majda and Qi J Stat Phys 2019, except that the total energy is one instead of pi.

@@

Then the Hamiltonian of \eqref{dimlessKdV1} is given by
\begin{equation}
\Ham = C_3 \drat^{-1} \, H_3 - C_2 \drat \, H_2
\end{equation}
where $\drat$ changes value across the ADC. More specifically, there exists an upstream and a downstream Hamiltonian
\begin{align}
&\Hup = C_3 \, H_3 - C_2 \, H_2 \\
&\Hdn = C_3 \dratdn^{-1} \, H_3 - C_2 \dratdn \, H_2
\end{align}
$\Hup$ is a conserved quantity for $t<0$ and $\Hdn$ is a conserved quantity for $t>0$. Each Hamiltonian $\Ham^{\pm}$ generates a corresponding Gibbs measure $\Gupdn$ 
\begin{align}
\Gupdn = C_{\thupdn} \exp(-\thupdn \Hupdn) \delta(\En - \pi)
\end{align}
where $\theta = \thupdn$ is the inverse temperature and $C_{\theta}$ a constant (the partition function) that depends on $\theta$. Each measure has corresponding ensemble mean $\mean{}_{\pm}$.

@@

The statistical matching condition is
\begin{equation}
\meanup{\Hdn} = \meandn{\Hdn}
\end{equation}
An experimental observation is that the incoming skewness is small, thus
\begin{equation}
\meanup{H_3} \approx 0
\end{equation}
which immediately gives
\begin{equation}
C_3 \dratdn^{-1} \meandn{H_3} = C_2 \dratdn \left( \meandn{H_2} - \meanup{H_2} \right)
\end{equation}
Simplifying gives
\begin{equation}
\label{H3H2}
\meandn{H_3} = \frac{C_2}{C_3} \dratdn^2 \,\, \Delta \mean{H_2}
\end{equation}
where $\Delta \mean{H_2} =  \meandn{H_2} - \meanup{H_2}$  indicates the difference between the upstream and downstream values.
To convert to dimensional, i.e.~experimental, values we note
\begin{align}
&\mean{u^3} = \frac{\mean{\eta^3}}{a^3} = 
\frac{\mean{\eta^3}}{\mean{\eta^2}^{3/2}} = \skw(\eta) \\
&\mean{u_x^2} = \left(\frac{\lamupdn}{a} \right)^2 \mean{\eta_x^2} 
= \left(\frac{\lamupdn}{a} \right)^2 \var(\eta_x)
\end{align}
Therefore
\begin{equation}
\meandn{H_3} = \frac{\pi}{3} \skw_+(\eta)
\end{equation}
and more subtle
\begin{equation}
\Delta \mean{H_2} =  \meandn{H_2} - \meanup{H_2} = 
\frac{\pi}{\amp^2} \left( \lamdn \var_+(\eta_x) - \lamup \var_{-}(\eta_x) \right) =
\frac{\pi}{\amp^2} \Delta \left( \lam \var(\eta_x) \right)
\end{equation}
Then \eqref{H3H2} gives the relationship
\begin{equation}
\frac{\skw(\eta)} {\Delta \left( \lam \var(\eta_x) \right) } =
\frac{3 C_2}{C_3} \amp^{-2} \dratdn^2 = \frac{\pi^2 \delup^2}{9 \lamfac^2} \
%3 \mu D^2 \left( \frac{h_0}{a} \right)^3 = 3 \eps^{-3} \mu D^2
\end{equation}
FINISH

In particular, the ratio on the left, which can be measured in the experiments, is predicted to scale as the inverse cube of the wave amplitude (as well as the square of the depth ratio).
